\vspace{-1.5em} \bf A general and efficient representation of ancestral recombination graphs

### 2024-04-18 paper.tex
Small tweaks to appendices

The schematic illustrates an important point about the biological reality of polytomies.
Three lineages coalesce in the left-hand genome of individual $D_{10}$, but do
so as the result of two successive bifurcations. This is \emph{necessarily} so,
because the only known method of reproducing DNA is by (semi-conservative) duplication.
Whether this polytomy is resolvable depends on the available mutational data.
Mutations can occur along any cell lineages. For example,
a mutation in the first cell division of $D_{10}$ could be shared between
the two gametes produced by the cells in the left half of $D_{10}$
but not shared by the right hand gamete.
With enough mutations, each
round of mitotic germline genome duplication within a single multicellular organism could
in principle be distinguished.

@@

% Moving on from ARG-unary to the more general case
More generally, locally unary nodes,
which can have one or more children in the graph,
are a common and important feature
of many different types of ARG.
As discussed in Appendix~\ref{sec-ARG-and-local-trees},
without these nodes marking the passage
of ancestral material through specific ancestors, the local trees
lack information about events other than local coalescence.
% Not sure if switching example will confuse people here, but would
% like to contrast with example where we don't use unary nodes to
% emphasise the point that we lose info
For example, the local trees for the classical event ARG
depicted in Fig.~\ref{fig-event-arg}B follow the usual conventions
and do not include any information about the recombination
that occurred at node \noderef{d}. Given these two local trees
in isolation
% (without further information about the corresponding ARG)
we lack specific information about the recombination.
Explicitly recording that node \noderef{d} lies on the
branch joining \noderef{b} to \noderef{e} in the left hand
tree, and \noderef{b} to \noderef{f} in the right hand tree
resolves all ambiguity, and makes the collection of local
trees exactly equivalent to the corresponding ARG (see previous section).
Unary nodes are a vital link between ARGs and local trees, and we
cannot fully reason about how a local tree is embedded in an ARG
without them. As we see in the next two sections, both ARG and locally
unary nodes occur in various scenarios, and are
produced by a range of current inference methods.

@@

The ARG in Fig.~\ref{fig-simplification}A is the output of a
backwards-time Wright-Fisher simulation for a sample of two diploid
individuals (population size $N=10$), and follows a similar process
to the methods described by~\cite{nelson2020accounting}.
As we proceed backwards in time, generation by generation, the
extant lineages choose parents randomly.
With a certain probability recombination occurs, and the ancestral
material of a lineage is split between the two parental
genomes. Local coalescence occurs
when lineages with overlapping ancestral material choose the same
parent genome.
Note that in this simulation we do not explicitly
model recombination \emph{events} via an ARG node, but simply record
the \emph{outcome} of a recombination via edges to
the parent's two genomes.
Thus, a recombinant node such as \noderef{g} in Fig.~\ref{fig-simplification}
may also correspond to a coalescence.
The distinction of
using a single node to represent
a recombination event, as is done in ~Fig.~\ref{fig-ancestry-resolution},
or two to represent the parent genomes, as in Fig.~\ref{fig-simplification},  is usually not important.
Either is possible in the gARG encoding, and the most convenient approach will vary by application
(discussed e.g.\ in Appendix~\ref{sec-big-and-little-arg}).
Note also that node \noderef{k} in Fig.~\ref{fig-simplification} has three children.
Polytomies like this are a natural feature of such a Wright-Fisher model
(but see Appendix~\ref{sec-cell-lineages-and-args}).

@@

An important consequence of simplifying ARGs to remove
unary nodes in local trees is that we lose some information
about recombination
events. This is related to the amount of \emph{precision} about
recombination events that we store and
can hope to infer from sampled genomes, which is the topic of
Appendix~\ref{sec-precision}.

>>>>

The schematic illustrates an important point about the biological reality of polytomies.
Three lineages coalesce in the left-hand genome of individual $D_{10}$, but do
so as the result of two successive bifurcations. This is necessarily so,
because the only known method of reproducing DNA is by (semi-conservative) duplication.
Whether this polytomy is resolvable depends on the available mutational data.
Mutations can occur along any cell lineages. For example,
a mutation in the first cell division of $D_{10}$ could be shared between
the two gametes produced by the cells in the left half of $D_{10}$
but not shared by the right hand gamete.
With enough mutations, each
round of mitotic germline genome duplication within a single multicellular organism could
in principle be distinguished.

@@

% Moving on from ARG-unary to the more general case
More generally, locally unary nodes,
which can have one or more children in the graph,
are a common and important feature
of many different types of ARG.
As discussed in Appendix~\ref{sec-ARG-and-local-trees},
without these nodes marking the passage
of ancestral material through specific ancestors, the local trees
lack information about events other than local coalescence.
% Not sure if switching example will confuse people here, but would
% like to contrast with example where we don't use unary nodes to
% emphasise the point that we lose info
For example, the local trees for the classical event ARG
depicted in Fig.~\ref{fig-event-arg}B follow the usual conventions
and do not include any information about the recombination
that occurred at node \noderef{d}. Given these two local trees
in isolation
% (without further information about the corresponding ARG)
we lack specific information about the recombination.
Explicitly recording that node \noderef{d} lies on the
branch joining \noderef{b} to \noderef{e} in the left hand
tree, and \noderef{b} to \noderef{f} in the right hand tree
resolves all ambiguity, and makes the collection of local
trees exactly equivalent to the corresponding ARG.
Unary nodes are a vital link between ARGs and local trees, and we
cannot fully reason about how a local tree is embedded in an ARG
without them.

@@

The ARG in Fig.~\ref{fig-simplification}A is the output of a
backwards-time Wright-Fisher simulation for a sample of two diploid
individuals (population size $N=10$), and follows a similar process
to the methods described by~\cite{nelson2020accounting}.
As we proceed backwards in time, generation by generation, the
extant lineages choose parents randomly.
With a certain probability recombination occurs, and the ancestral
material of a lineage is split between the two parental
genomes. Local coalescence occurs
when lineages with overlapping ancestral material choose the same
parent genome.
Note that in this simulation we do not explicitly
model recombination \emph{events} via an ARG node, but simply record
the \emph{outcome} of a recombination via edges to
the parent's two genomes.
Thus, a recombinant node such as \noderef{g} in Fig.~\ref{fig-simplification}
may also correspond to a coalescence.
The distinction of
using a single node to represent
a recombination event, as is done in ~Fig.~\ref{fig-ancestry-resolution},
or two to represent the parent genomes, as in Fig.~\ref{fig-simplification},  
is often not important.
Either is possible in the gARG encoding, and 
the most convenient approach will vary by application.
Note also that node \noderef{k} in Fig.~\ref{fig-simplification} has three children.
Polytomies like this are a natural feature of such a Wright-Fisher model.
See Appendix~\ref{sec-cell-lineages-and-args} for a discussion of the 
ultimate biological interpretation of these topological considerations.

@@
